Do moral communities have a spatial dimension? A spatial exploratory analysis of places of worship and violent crime in the city of Recife, Brazil

### 2020-06-11 Moral-Communities-and-Crime/Response-to-Reviewers-v2.Rmd

---
title: "Response to Reviewers"
output: pdf_document
---

We would like to thank the editor and two anonymous reviewers for their thoughtful comments and suggestions to revise the paper. We believe that the paper has been substantially improved by their input. In this letter we respond to the feedback received, and describe the actions taken to revise the paper.

@@

>>\textcolor{blue}{2. Several types of community organizations, including schools, clubs, and neighborhood associations, can strengthen the capabilities of communities to exert informal social control.}

>>\textcolor{blue}{3. Of several types of community organizations, churches are among the most overtly moral, since they often have prescriptive codes of behavior.}

>>\textcolor{blue}{4. Why would the location of places of worship matter? Groff (2015) reviews several theoretical perspectives that address the role of informal social control in micro-places: these include "eyes on the street" and human territorial functioning. By affecting informal surveillance, movement governing rules, direct intervention, and the social climate at places, places of worship are hypothesized to contribute to informal social controls.}

>\textcolor{blue}{We believe that this argument is logical and sound, and churches, in fact, have been studied in the past for their potential to suppress crime (Abdullah, 2018; Davignon, 2015; Furr, 2010; Traunmuller, 2011). A more novel aspect is the study of places of worship, as distinct entities from the members of the congregations: as Warner and Konkel (2019) note, this is an aspect that has received less attention in empirical and theoretical research. }

>\textcolor{blue}{The second question is related to the above: if places of worship somehow help to strengthen social control, does this happen at the level of points or for areas? In our view, Groff (2015) already explains why these investigations should matter at the micro-level. But there is also an analytical aspect to this question: what is an appropriate unit of support for the analysis? In this sense, we would note that aggregating point data to areas inevitably involves loss of information. Sometimes this situation is unavoidable (say, statistics are recorded for areas, so locational information is lost from the origin.) But when we have access to detailed geographical information, converting to area data is an unnecessary loss of information. The question that we are asking is actually quite geographical: if there is evidence of a moral community effect, what is its distance decay? This question is best answered by considering the distance from criminal events to places of worship, instead of simple propinquity of criminal events and places of worship in areas.}

@@

>\textcolor{blue}{We have edited the paper to increase its clarity.}

The current framing of the article also does not convincingly situate 'moral communities' within a geographical framework and does not make clear the hypothesized mechanisms that link the presence/absence of a church at a specific location with criminal behavior and, more specifically, the presence/absence of churches of specific religions with violent crimes. The authors should make sure not to conflate the processes linking crime with population characteristics (percent of individuals who are religious in an area), with religion more broadly (hellfire hypothesis), and with the location of a church in geographic space. This theoretical work is necessary to justify why this study is relevant and make clear how it contributes to the existing literature.

>\textcolor{blue}{Thank you for this comment. As pointed out above, the theoretical work was discussed by Groff, among others. Although conceptually the study draws from much earlier research (e.g., Appleyard, Jacobs, Taylor), empirical work has lagged, as recently noted by Warner and Konkel (2019). This is where the current paper aims to make a contribution.}

@@

The introduction of the paper does a really good job setting out the importance of reducing violent crime for achieving policy goals, but it is not clear how this research will contribute to developing, implementing, or evaluating policy. For this journal, I think it is necessary for the research to contribute to policy; so, how can police, religious officials, and other policymakers use the results of this study?

>\textcolor{blue}{The analysis of elements that can influence the occurrence of the criminal act is able to increase the range of tools that public authorities can use to face the problem, particularly in a reality as difficult as that of Brazil. The notion of informal social control allows new strategies to combat crime to be explored. In particular, we find that Catholic churches do not "project" any more moral community than pharmacies, which means that there is probably not much that can be done in terms of leveraging them for informal social control purposes. Spiritist churches have a lower relative risk at relatively short distances of less than approximately 50 m (which would be the side of a small block). Why is this? We cannot posit a causality relationship, but could this be due to superstitious beliefs associated by laypeople to this religion? And if so, could this be leveraged for policy purposes? Evangelical churches, on the other hand, have a higher relative risk than any other establishment or place of worship at distance between 20-100 m. A relevant question is whether this is crime by Evangelicals or crime towards Evangelicals. This is not something that the analysis can throw light on, but that deserves further investigation to determine whether it can lead to actionable policy recommendations.}

@@

I currently find the logic of the analysis and the results to be confusing and I think that adding the results of more exploratory work before showing results of very specific analyses (i.e., those with SES quintiles and different church types) would help to provide a convincing argument that the results of this research are solid. For example, the authors may first examine the intensity of crime as it relates to all churches (aggregated), then to adjust for ses with all churches, and then to adjust for land use. Then, after presenting the results of the initial analysis for the aggregated churches, the results of the specific church types can be presented.

>\textcolor{blue}{We hope that the revisions elsewhere (e.g., an improved presentation of the argument behind the analysis) have resolved this issue.}

@@

The authors should add a review of social disorganization theory and informal social control as these concepts are mentioned a number of times but not discussed in any detail. The similarities and differences between these concepts and the geographically-situated moral community concept should also be discussed.

>\textcolor{blue}{The referee's suggestion has been incorporated into the text. Two footnotes have been included that detail theories of social disorganization and the theory of social control. The focus was to present the main characteristics of these theoretical views and to relate them to the geographical aspect. In general, both views highlight the relationship between local characteristics (neighborhoods, for example), social networks and criminal behavior.}

The authors should justify their choice of method over other approaches, such as regression models, which provide a framework for quantifying the relationship between crime and churches after accounting for SES factors, land use types, and population at risk. There are a number of regression models that allow  coefficients to vary over space, which would quantify how the impact of churches on crime changes by distance (see: generalized additive models, spatially-varying coefficients, geographically weighted regression). Without this type of analysis, or at least a more thorough discussion of why this point pattern method was chosen, I am not convinced that crime is, in fact, associated with evangelical churches at close distances and that this isn't due to other factors, like evangelical churches being located in high population density areas (and so crime is due to higher population density, not the presence of a church), for example.

>\textcolor{blue}{Thank you for this comment. We agree, there are many methods that could be applied. In our analysis we use three tools:}

>\textcolor{blue}{1. Quadrat counts, which are discrete estimates of the intensity of a point process in a region.}

>\textcolor{blue}{1. The relative distribution, which is a smooth estimate of the intensity of a process as a function of distance to a signal event.}

@@

>\textcolor{blue}{2. The relative distribution function with a baseline function, which adjusts the relative distribution using other risk factors. In our case, those other factors are summarized by the SED score. The interpretation of the relative distribution with a baseline is as follows: values of rho greater than 1 indicate that the intensity of crime is higher over what the baseline explains, and values less than 1 indicate that the intensity of crime is lower than what the baseline explains.}

>\textcolor{blue}{In the previous version of the paper, our SED score was calculated using several potential confounders of crime: median income, proportion of population unemployed, proportion of households with a single mother, proportion of households with a young single mother, proportion of population in poverty. In the current version of the paper, we also include population density in the calculation of the SED score (thank you for the suggestion). Accordingly, the relative distribution with a baseline function is the smooth estimate of the intensity of crime after controlling for potential confounders, including population density.}

>\textcolor{blue}{We believe that this approach is informative, and since we are interested in the distribution of crime in the proximity of places of worship, it is also appropriate. Furthermore, by considering other underlying risk factors (in the form of the SED score), we believe that we have diligently covered most reasonable confounders. We have edited the text to explain this.}

@@

Did this study investigate the location of crime events, offender locations, or victim locations? I assume crime events, but it would be good to clarify nonetheless.

>\textcolor{blue}{In the data section, we changed the description. In the third paragraph, we use the term crime event location, which is more precise.}

For the intensity measures, what is the unit area for the results? Would it be possible to provide an intensity unit such as per mile^2 or per km^2, for example? I think that this would improve the interpretation of the results and Figures 6 and 7 in particular.

>\textcolor{blue}{Thank you for this comment. The intensity is in events/square meter, except in the case of the relative distribution with a baseline function, which is dimensionless (see below).}

When discussing the quantitative results throughout Sections 5.3 and 5.4, please add uncertainty intervals so that readers can interpret the degree to which two quantities are, in fact, different? For example, p.16-line 49: "reaching a peak of 2.4916 * 10e-5 (95% CI: 2.1 * 10e-5, 3.2 * 10e-5)."

>\textcolor{blue}{Thank you for this comment. The intervals of confidence are now reported in the figures.}

Please provide some additional text describing the equations in Section 3.2. It is not immediately clear to me how to interpret equations and how they are reflected in the results. For example, what is B(s), how is it calculated, and what does it mean for understanding the distribution of crime in relation to churches?

>\textcolor{blue}{Both quadrats and the relative distribution function are estimators of intensity. In the case of quadrats, the estimator is the number of events in the quadrat divided by area of the quadrat (conditional quadrat counts use irregular quadrats that are defined based on some control variable, in our case SED). In the case of the relative distribution, the output is an estimate of intensity conditioned on distance from a signal event, i.e., a place of worship. The relative distribution can be adjusted by means of a baseline function B(s), which is also an estimate of intensity but conditioned on some control covariate (in our cases SED). Therefore, the interpretation of the relative distribution with a baseline function is as a ratio of the intensity with respect to distance to the baseline intensity according to the control covariate. If the ratio, which is dimensionless, is 1, then intensity as a function of distance is equal to the baseline intensity.}

@@

For readers (like me) who are unfamiliar with the geography of Recife, Brazil, it would be helpful if a detailed map was included in the paper that shows where the downtown is, major roads, different neighborhoods, etc. This could be included in the background of Figure 1 or in an Appendix. It would also be helpful if the authors provided some description of the patterns observed in Figure 1 and Figure 2 identifying where violent crimes (or churches, supermarkets, etc.) were clustered and where they weren't clustered.

>\textcolor{blue}{We now include a figure with maps for socio-economic and demographic variables which help to provide context to SED. We also have included a brief explanation of the requested clustering pattern in the text and recreated the figures after adding some transparecy to compensate for overplotting. This should make it easier to appreciate the patterns.}

@@

>\textcolor{blue}{Thank you for reading the paper and for your thoughtful suggestions to improve it.}

We’ll start with a couple of minor issues

– in the quadrat analysis are the results conditioned on the quadrat size. If I understand correctly this might not be an issue with conditional quadrat counts, through the use of a covariate.  It does remind me of Openshaw early work on optimal zoning systems.  Perhaps a simple example to guide the reader might help.

>\textcolor{blue}{The results are conditioned by the size of the quadrat, but in this case since the quadrats are defined based on a covariate, we have a conceptual reason to prefer them to other partitions (each quintile in the SED is one quadrat). We have explained in the text how the quadrats function as containers for counting events.}

@@

- the symbolism in Figure 1, 2, and 3 are rather large.  I’m at a loss to interpret Figure 3, because as the image is built up (I have a rather slow graphics card) I see symbols overlapping.

>\textcolor{blue}{Thank you for this comment. We have recreated the figures with some transparency to compensate for overplotting.}

@@

There’s a cultural issue which I find challenging.  I think I know what Catholic and Evangelical churches may be, but this is first encounter with Spiritism. I assume that the first two are differentiated along some transubstantiation/consubstantiation divide, with the Catholics being in the former group. As for Spiritist churches I’m at a loss.  Do these follow the teaching of Allan Kardec; I presume it’s related to the Spiritualist churches found in the Anglophone countries, which go back to Swedenborg and Mesmer, and which seem to be originally motivated by trying to make contact with the dead. If so, then it’s very much a minority sect.  Perhaps the authors might like to expand on the details further for those readers who are not familiar with religious practices, and how they are undertaken in Brazil. Would I recognise the same in Cape Town, to take a city at random. Would this be a typical Spiritist church: https://goo.gl/maps/kwWe2B7didMk8vfDA

>\textcolor{blue}{We have included in the text some additional information of the religions studied in the article.}

@@

Some description of the geography of Recife would also help. Where are residential areas, where are industrial zones, is there a university, and so on.

>\textcolor{blue}{The appendix included information that allows a more in-depth characterization of the city of Recife: population distribution, income, inequality, urban facilities (education and health) and location of residential areas.}

What’s the reason for the choice of controls – what if we take another three, such as take-away food outlets, garden centres, and clothing/footwear shops? Would we get similar of different results? You could use simulation to build a distribution of randomly chosen groups of different land uses. I’m not proposing that this should happen, but it’s a possibility.

>\textcolor{blue}{The results for the various controls are qualitatively similar both for quadrat analysis and relative distribution. After choosing these three types of establishments semi-randomly and verifying their similarities, we think it unlikely that other types of common establishments would change the results. In this case simulation of a random variable is not useful, because it would detect clustering of all the events. Rather, what we are interested in is clustering of one type of event with respect to another.}

@@

I am reminded of Openshaw’s early work on cluster detection – work was based around trying to link clusters of leukaemia to electricity substations, and overhead transmission lines. Alan Craft was rumoured to have asked something along the lines of “how do we know it’s substations, or fish chip shops.  What if it’s telephone booths?”  Openshaw turned his approach on its head to ask where the clusters were occurring. This lead him to a nuclear waste reprocessing plant, and an urban domestic waste incineration plant. Does the possibility of turning the question to look at this from another point of view?

>\textcolor{blue}{This is a very perceptive comment. In fact, the approach that Openshaw pioneered was later converted into several forms of spatial scan statistics. Its legacy is still felt here: we can explore co-location patterns by considering signal events that we are interested in for theoretical/conceptual reasons (i.e., places of worship).}

@@

I’m left wondering whether this paper tells me something about spatial relationships in Recife.  The author(s) assert that the research is exploratory.  But to what extent might they apply elsewhere? Would these relationships translate to Vienna, for example?

>\textcolor{blue}{The specific results are applicable to Recife only. However, the general principle could be more general, and it aligns well with one to two ideas: the Violence of the Just, or the possibility that violence is a consequence of religious intolerance. Irrespective of which of these two propositions is true, the general principle could be in operation in areas where conservative Evangelicals are present (similar results, although at the aggregated level, have been reported by Desmond et al., 2010). For this reason, Vienna might be a stretch, but other cities in Latin America and North America not so much.}

@@

I think that ASAP readers might be find this paper rarified.  They certainly need help with some of the cultural issues.

>\textcolor{blue}{We have made an effort to provide additional context, which we hope will make the paper more approachable to readers who are less familiar with the Latin American context.}

### 2020-06-11 Moral-Communities-and-Crime/Moral_Communities_and_Crime.Rmd
v2 Finished revisions, sent back to Edivaldo and Tatiane for comments.

# Load packages used in this document

library(tidyverse)
library(kableExtra)
library(sf)
library(spatstat)
library(maptools)
library(ggthemes)
library(raster)
library(fasterize)
```

@@

2. If so, what is the distance decay of said moral community effect? Is it uniform over a certain distance or does it vary?

The case study is the city of Recife in the state of Pernambuco, in Brazil's Northeast. Recife is a large and important metropolitan area in a historically poor region, and afflicted by high levels of violent crime, having the dubious honor of being one of five state capitals with the highest rates of homicide in the period under study [@Menezes2013spatial]. The empirical strategy is to use spatial analysis to explore the potential geographical relationships between violent criminal events, on the one hand, and places of worship and a selection of commercial establishments that serve as controls, on the other. Dissagregated data allows us to analyze the relevant point patterns at a very high level of resolution.

After this introductory section, the rest of this paper is structured as follows: Section \ref{background} reviews some relevant theoretical perspectives on religion and crime; then Section \ref{methods} presents the empirical strategy used in the work. Section \ref{case} presents the context as well as the data used in the study; Section \ref{results} presents the analysis, whereas the results are discussed in Section \ref{discussion}. Finally, some concluding remarks are offered in Section \ref{conclusions}.

@@

The hypothesis of moral communities has over the years been used to examine a variety of outcomes of interest. Recent examples include Stroope and Baker's [-@Stroope2018moral] exploration of religiosity and self-rated health; Davignon and Thomson [-@Davignon2015christian] with their research on institutional context and the religiosity of students; and the study of religion as a source of trust of Traunmuller [-@Traunmuller2011moral]. This is in addition, of course, to numerous studies on criminal behavior such as Eitle [-@Eitle2011religion], an author who explored the deterrence power of religion on gambling; Lee and Bartkowski's [-@Lee2004love] investigation of juvenile homicide in rural areas; and the research of Regnerus [-@Regnerus2003moral] on adolescent delinquency. The hypothesis of moral communities as a form of informal social control, on the other hand, has been less studied from a geographical perspective, and it is only recently that has attracted the attention of researchers. Groff [-@Groff2015informal], for instance, discusses informal social control as a phenomenon that can plausibly operate at different geographical scales, from the level of the home and family, through the street block and neighborhood, possibly to other scales such as the county. In this way, Nie and Yang [-@Nie2019smoking] remark on a recent paper on the lack of research conducted to study how the religious context of a geographical area (e.g., a county) may influence (youth smoking) behavior (p. 2). This point is echoed by Warner and Konkel [-@Warner2019neighborhood], who moreover recommend the use of even smaller units of analysis (e.g., Census Block Groups), in an effort to reduce aggregation bias [see @Hipp2007block], but more importantly because the processes defined by social disorganization theory\footnote{The social disorganization theory (Shaw e McKay, 1942)  presents the social structure as an indispensable factor for the elaboration of theoretical and methodological models related to crime in urban neighborhoods. According to Villareal e Silva [-@Villarreal2006social], “neighborhoods characterized by dense social networks will experience greater trust among residents and cooperation in the enforcement of social norms against crime. Disadvantaged neighborhoods are thought to have fewer social ties among residents and lower rates of organizational participation”. The argument is that community characteristics such as poverty, ethnic heterogeneity and residential mobility disrupt the social organization of urban neighborhoods in such a way as to reduce their capacity to exercise social control [@Rodrigues2012medo]. Thus, aspects such as residential instability or low socioeconomic status would contribute to the weakening of the community organization, increasing crime rates. This theory highlights the importance of the geography of crime and the characteristics of disorder in its analysis.} are thought to occur at relatively small geographical scales.

The preceding review, as well as other research [-@Malleson2019identifying], makes clear the need for research at various geographical scales. From a theoretical perspective, the social mechanisms that underpin the hypothesis of moral communities and informal social control can happen at various geographical scales, some of which previous research has already addressed: the work of Nie and Yang [-@Nie2019smoking], for one, highlights the role of processes detectable at a relatively high level of aggregation, whereas Warner and Konkel [-@Warner2019neighborhood], looking specifically at the effect of places of worship, bring the study closer to the level of the neighborhood. In what follows, we propose to adopt a much more disaggregated approach by considering the intensity of criminal events with respect to places of worship. This requires micro-level data both of crimes and places of worship, and the use of appropriate spatial analytical tools for point patterns, as discussed next.

@@

Brazil is a country with a Christian predominance. In the first national census in which religion was considered, in 1872, practically the entire population (99.72%) identified themselves as Catholic. In the 1940 census, when the religion question was researched again, Catholicism was still the religion with which the vast majority of the population identified: 95.2% of Brazilians declared themselves as Catholics that year.

The dominance of Catholicism in the religious life of Brazil is unsurprising given the colonial past of the country. However, the religious makeup of the population started was not completely homogeneous. Among other offerings in the market place of spiritual ideas, Spiritism stands out. The Spiritist Doctrine was founded in nineteenth-century France by Allan Kardec as a way of understanding the world and its relations with the "beyond": a curious mix of pseudo-scientific language, philosophy, and religion [@Arribas2011doutrina]. Spiritism arrived in Brazil in the middle of the previous century and since then it slowly spread in large and medium-sized cities. Today, Spiritism is the third largest religious segment in Brazil, after Catholicism and Evangelical Christianism, and shares with these religions the Christian principle of charity through pious assistance.

Another signal event in the developmento of religion in Brazil is the arrival of Pentecostal Protestant churches, such as the Christian Congregation of Brazil in 1910 and the Assembly of God in 1911. Initially, these religious denominations did not make important inroads in Brazilian society (as evidenced by the 1940 census). However, in the context of rapid industrialization and major urbanization, they contributed to the change in the religious landscape along with other religions [@deSouza2019pluralidade]: according to the 2010 census the configuration of the religious scene in Brazil had changed considerably from the previous century: 64.6% of the population identified as Catholic and 22.2% as Evangelical Christians. The acceleration of this demographic change was directly related to the inroads made by Evangelical churches: The Assembly of God, for instance, grew considerably to total more than 12.3 million adherents. This church is represented extensively in many regions of the country, following a system of loosely allied ministries. With a charismatic profile and the lack of a centralized authority, growth of this church has occurred mainly through subdivision into numerous small and autonomous churches. These churches have in common the Prosperity Gospel and an emphasis on economic entrepreneurship  [e.g., @Mora2008marketing]. They also do not shy away from partisan political action.

In general, there have been two major changes in Brazil regarding religious issues: the slow decline of Catholicism and the rapid growth of Evangelical Christianism. These changes are evident in the daily life of the average Brazilian, either through the radio or through TV programs, which give more and more space to religious doctrines. At the same time, there has been an increase in cases of religious intolerance in Brazil, generating violence acts against temples, objects of worship, and people.

@@

where $\lambda(s)$ is the intensity of the process at point $s$, given by the expected number of events in a small area $\text{d}s$ around $s$, as the area becomes arbitrarily small. If the process is homogeneous (i.e., the intensity is a constant over space), an apt estimator of the intensity is the global intensity, simply the number of events divided by the area of the region under analysis:
$$
\hat{\lambda} = \frac{n}{|\mathcal{A}|}
$$

@@

@@

ggplot(data = events_recife.sf %>% filter(Type == "Violent Crime")) + 
  geom_sf(color = "red") +
  geom_sf(data = recife.sf, fill = NA) + 
  theme_tufte()
```

@@

ggplot(data = events_recife.sf %>% filter(Type != "Violent Crime")) + 
  geom_sf(aes(color = Type, fill = Type)) +
  geom_sf(data = recife.sf, fill = NA) + 
  facet_wrap(~ Type, nrow = 2) +
  scale_fill_viridis_d(option = "viridis", direction = 1) +
  scale_color_viridis_d(option = "viridis", direction = 1) +
  theme_tufte()  +
  theme(axis.text.x = element_text(angle = 90))

@@

The analysis is implemented using the `R` statistical computing language, and documents to replicate the analysis, in addition to all data necessary, can be obtained from this anonymous shared folder:

https://drive.google.com/open?id=1tuJM4Mhi0Ftq3ZEwjv6RP9vO3veGjIR_

@@

As seen in the table, Recife is a city with large socio-economic and demographic disparities; for example, the _setor_ with the highest median income has a median income that is `r prettyNum(filter(sed_descriptives, variable == "Median Income")["max"]/filter(sed_descriptives, variable == "Median Income")["min"] * 100, digits = 2, big.mark = ",")`% higher than the median income in the _setor_ with the lowest median income. Whereas there are _setores_ where the proportion of population who are unemployed is zero, there are _setores_ where almost 60% of the population are unemployed. Likewise, there are _setores_ where almost 70% of the population lives in poverty. In addition to these economic indicators, two variables are used to represent family disruption, the proportion of families whose head is a single mother, and the proportion of families whose head is a _young_ single mother, that is, a woman between the ages of 15 and 25. As can be seen, there are _setores_ with approximately 70% of households led by single mothers, and of these, over 21% are led by younger women, indicating a high degree of family disruption.

For the analysis, we use Principal Component Analysis, a data reduction technique, to obtain an indicator of Socio-Economic Deprivation, essentially the first principal component in the output. The loadings of this factor (Table \ref{tab:sed-descriptive-statistics}) indicate that high Socio-Economic Deprivation is a combination of (in terms of importance): high levels of poverty, high levels of unemployment, and low median income, followed by high levels of family disruption, in particular proportion of families led by young single mothers. This factor accounts for almost 54% of the variance..

@@

>>>>

# Load packages used in this document
library(fasterize)
library(ggthemes)
library(gridExtra)
library(kableExtra)
library(maptools)
library(raster)
library(sf)
library(spatstat)
library(tidyverse)
```

@@

2. If so, what is the distance decay of said moral community effect? Is it uniform over a certain distance, such that it could be well approximated by areas, or does it vary?

The case study is the city of Recife in the state of Pernambuco, in Brazil's Northeast. Recife is a large and important metropolitan area in a historically poor region, and afflicted by high levels of violent crime, having the dubious honor of being one of five state capitals with the highest rates of homicide in the period under study [@Menezes2013spatial]. The empirical strategy is to use spatial analysis to explore the potential geographical relationships between violent criminal events, on the one hand, and places of worship and a selection of commercial establishments that serve as controls, on the other.

Note that this is a reproducible research document. The analysis is implemented using the `R` statistical computing language, and documents to replicate the analysis, in addition to all data necessary, are available from a repository\footnote{\url{https://drive.google.com/open?id=1tuJM4Mhi0Ftq3ZEwjv6RP9vO3veGjIR_}}.

@@

The hypothesis of moral communities has over the years been used to examine a variety of outcomes of interest. Recent examples include Stroope and Baker's [-@Stroope2018moral] exploration of religiosity and self-rated health; Davignon and Thomson [-@Davignon2015christian] with their research on institutional context and the religiosity of students; and the study of religion as a source of trust of Traunmuller [-@Traunmuller2011moral]. This is in addition, of course, to numerous studies on criminal behavior such as Eitle [-@Eitle2011religion], an author who explored the deterrence power of religion on gambling; Lee and Bartkowski's [-@Lee2004love] investigation of juvenile homicide in rural areas; and the research of Regnerus [-@Regnerus2003moral] on adolescent delinquency. The hypothesis of moral communities as a form of informal social control, on the other hand, has been less studied from a geographical perspective, and it is only recently that has attracted the attention of researchers. Groff [-@Groff2015informal], for instance, discusses informal social control as a phenomenon that can plausibly operate at different geographical scales, from the level of the home and family, through the street block and neighborhood, possibly to other scales such as the county. In this way, Nie and Yang [-@Nie2019smoking] remark on a recent paper on the lack of research conducted to study how the religious context of a geographical area (e.g., a county) may influence (youth smoking) behavior (p. 2). This point is echoed by Warner and Konkel [-@Warner2019neighborhood], who moreover recommend the use of smaller units of analysis (e.g., Census Block Groups), in an effort to reduce aggregation bias [see @Hipp2007block], but more importantly because the processes defined by social disorganization theory\footnote{The social disorganization theory (Shaw e McKay, 1942)  presents the social structure as an indispensable factor for the elaboration of theoretical and methodological models related to crime in urban neighborhoods. According to Villareal e Silva [-@Villarreal2006social], “neighborhoods characterized by dense social networks will experience greater trust among residents and cooperation in the enforcement of social norms against crime. Disadvantaged neighborhoods are thought to have fewer social ties among residents and lower rates of organizational participation”. The argument is that community characteristics such as poverty, ethnic heterogeneity and residential mobility disrupt the social organization of urban neighborhoods in such a way as to reduce their capacity to exercise social control [@Rodrigues2012medo]. Thus, aspects such as residential instability or low socioeconomic status would contribute to the weakening of the community organization, increasing crime rates. This theory highlights the importance of the geography of crime and the characteristics of disorder in its analysis.} are thought to occur at relatively small geographical scales.

Groff [@Groff2015informal], in particular, explains the small-scale mechanisms that could be in operation in terms of "Eyes on the Street" and Human Territorial Functioning. "Eyes on the Street" [see @Jacobs1961death; @Jacobs1968community; @Appleyard1980livable] refers to involvment in small civic acts by users of public spaces, which signal social organization to an observer. Community engagement, moreover, is engendered at small scales, as shown by authors that include Appleyard [-@Appleyard1980livable] and more recently Grannis [-@Grannis2009from]. Human Territorial Functioning [HTF; @Taylor1988human; @Taylor1998crime] also provides a valuable perspective on social informal control This concept posits that territorial functioning is a product of the shared perceptions of individuals in regards to places, and how these perceptions are shaped by the physical, social, and cultural environments. For example, Paez [-@Paez2013mapping] discusses how perceptions of safety are the outcome of a socially interactive process, and shows how this process leads to spatial patterning at very small-scales. The preceding review, as well as other research [-@Malleson2019identifying], makes clear the need for research at various geographical scales. From a theoretical perspective, the social mechanisms that underpin the hypothesis of moral communities and informal social control can happen at various geographical scales, some of which previous research has already addressed: the work of Nie and Yang [-@Nie2019smoking], for one, highlights the role of processes detectable at a relatively high level of aggregation, whereas Warner and Konkel [-@Warner2019neighborhood], looking specifically at the effect of places of worship, bring the study closer to the level of the neighborhood.

@@

Brazil is a predominantly Christian country. The first national census that collected information on religion was in 1872, a year when practically the entire population (99.72%) identified themselves as Catholic. In the 1940 census, when the religion question was asked again, Catholicism was still the dominant religion with 95.2% of Brazilians declaring themselves as Catholics. The dominance of Catholicism in the religious life of Brazil is unsurprising given the country's past as a former colony of a Catholic power (Portugal). However, the religious makeup of the population started to become less homogeneous in the second half of the twentieth century. Among other offerings in the market place of spiritual ideas, Spiritism stands out. The Spiritist Doctrine was founded in nineteenth-century France by Allan Kardec as a way of understanding the world and its relations with the "beyond", in a curious mix of pseudo-scientific language, philosophy, and religion [@Arribas2011doutrina]. Spiritism arrived in Brazil in the middle of the nineteenth century, and since then it slowly spread in large and medium-sized cities. Today, Spiritism is the third largest religious segment in Brazil, after Catholicism and Evangelical Christianism, and shares with these religions the Christian principle of charity through pious assistance.

Another signal event in the developmento of religion in Brazil was the arrival of Pentecostal Protestant churches, such as the Christian Congregation of Brazil in 1910 and the Assembly of God in 1911. Initially, these religious denominations did not make important inroads in Brazilian society (as evidenced by the 1940 census). However, in the context of rapid industrialization and major urbanization, they contributed to the change in the religious landscape along with other religions [@deSouza2019pluralidade]: according to the 2010 census the configuration of the religious scene in Brazil had changed considerably from the previous century: 64.6% of the population identified as Catholic and 22.2% as Evangelical Christians. The acceleration of this demographic change was directly related to successful proselitism by Evangelical churches: The Assembly of God, for instance, is a church that grew considerably to total more than 12.3 million adherents. This church is represented extensively in many regions of the country, following a system of loosely allied ministries. With a charismatic profile and the lack of a centralized authority, growth of this church has occurred mainly through subdivision into numerous small and autonomous churches. These churches have in common the Prosperity Gospel and an emphasis on economic entrepreneurship  [e.g., @Mora2008marketing]. They also do not shy away from partisan political action.

In general, there have been two major changes in Brazil regarding religious issues: the slow decline of Catholicism and the rapid growth of Evangelical Christianism. These changes are evident in the daily life of the average Brazilian, either through the radio or through TV programs, which give more and more space to religious doctrines. At the same time, there has been an increase in cases of religious intolerance in Brazil, generating violence acts against temples, objects of worship, and people [e.g., @deSouza2019pluralidade; @Fonseca2018primeiras].

In what follows, we propose to adopt a disaggregated approach by considering the intensity of criminal events with respect to places of worship. This requires micro-level data both of crimes and places of worship, and the use of appropriate spatial analytical tools for point patterns, as discussed next.

@@

\noindent where $\lambda(s)$ is the intensity of the process at point $s$, given by the expected number of events in a small area $\text{d}s$ around $s$, as the area becomes arbitrarily small. If the process is homogeneous (i.e., the intensity is a constant over space), an apt estimator of the intensity is the global intensity, simply the number of events divided by the area of the region under analysis:
$$
\hat{\lambda} = \frac{n}{|\mathcal{A}|}
$$

@@

## Interpretation

Both quadrats and the relative distribution function are estimators of intensity. In the case of quadrats, the estimator is the number of events in the quadrat divided by area of the quadrat (conditional quadrat counts use irregular quadrats that are defined based on some control variable, in our case SED). In the case of the relative distribution, the output is an estimate of intensity conditioned on distance from a signal event, e.g., a place of worship. The relative distribution can be adjusted by means of a baseline function $B(s)$, which is also an estimate of intensity but conditioned on some control covariate (in our cases SED). Therefore, the interpretation of the relative distribution with a baseline function is as a ratio of the intensity with respect to distance, to the baseline intensity according to the control covariate. If the ratio, which is dimensionless, is 1, then intensity as a function of distance is equal to the baseline intensity.

@@

ggplot(data = events_recife.sf %>% filter(Type == "Violent Crime")) + 
  geom_sf(color = "red", alpha = 0.1, shape = 16) +
  geom_sf(data = recife.sf, fill = NA) + 
  theme_tufte()
```

@@

ggplot(data = events_recife.sf %>% filter(Type != "Violent Crime")) + 
  geom_sf(aes(color = Type, fill = Type), alpha = 0.1, shape = 16) +
  geom_sf(data = recife.sf, fill = NA) + 
  facet_wrap(~ Type, nrow = 2) +
  scale_fill_viridis_d(option = "viridis", direction = 1) +
  scale_color_viridis_d(option = "viridis", direction = 1) +
  theme_tufte()  +
  theme(axis.text.x = element_text(angle = 90))

@@

@@

@@

As seen in the table, Recife is a city with large socio-economic and demographic disparities; for example, the _setor_ with the highest median income has a median income that is `r prettyNum(filter(sed_descriptives, variable == "Median Income")["max"]/filter(sed_descriptives, variable == "Median Income")["min"] * 100, digits = 2, big.mark = ",")`% higher than the median income in the _setor_ with the lowest median income. Whereas there are _setores_ where the proportion of population who are unemployed is zero, there are _setores_ where almost 60% of the population are unemployed. Likewise, there are _setores_ where almost 70% of the population lives in poverty. In addition to these economic indicators, two variables are used to represent family disruption, the proportion of families whose head is a single mother, and the proportion of families whose head is a _young_ single mother, that is, a woman between the ages of 15 and 25. As can be seen, there are _setores_ with approximately 70% of households led by single mothers, and of these, over 21% are led by younger women, indicating a high degree of family disruption. Unlike other places, where higher population density is associated with more poverty, the distribution of this variable in Recife is more mixed: middle income and higher income households favor high density development in many parts of the city. The spatial distribution of these variables can be seen in Figure \ref{fig:plot-sed-variables}.

```{r fig-plot-sed-variables, echo=FALSE, fig.height= 9, fig.align = "left", fig.cap="\\label{fig:plot-sed-variables}Spatial distribution of socio-economic and demographic variables in Recife"}

median_income.plot <- ggplot(setores.sf) + 
  geom_sf(aes(fill = median_income), color = NA) +
  geom_sf(data = recife.sf, fill = NA) + 
  scale_fill_viridis_c(name = "Median Income", option = "viridis", direction = -1) +
  theme_tufte()  +
  theme(axis.text.x = element_text(angle = 90),
        legend.position = "bottom")

prop_unemployed.plot <- ggplot(setores.sf) + 
  geom_sf(aes(fill = prop_unemployed), color = NA) +
  geom_sf(data = recife.sf, fill = NA) + 
  scale_fill_viridis_c(name = "Prop. Unemployed", option = "viridis", direction = -1) +
  theme_tufte()  +
  theme(axis.text.x = element_text(angle = 90),
        legend.position = "bottom")

prop_poverty.plot <- ggplot(setores.sf) + 
  geom_sf(aes(fill = prop_poverty), color = NA) +
  geom_sf(data = recife.sf, fill = NA) + 
  scale_fill_viridis_c(name = "Prop. Poverty", option = "viridis", direction = -1) +
  theme_tufte()  +
  theme(axis.text.x = element_text(angle = 90),
        legend.position = "bottom")

prop_unemployed.plot <- ggplot(setores.sf) + 
  geom_sf(aes(fill = prop_unemployed), color = NA) +
  geom_sf(data = recife.sf, fill = NA) + 
  scale_fill_viridis_c(name = "Prop. Unemployed", option = "viridis", direction = -1) +
  theme_tufte()  +
  theme(axis.text.x = element_text(angle = 90),
        legend.position = "bottom")

prop_single_mother.plot <- ggplot(setores.sf) + 
  geom_sf(aes(fill = prop_single_mother), color = NA) +
  geom_sf(data = recife.sf, fill = NA) + 
  scale_fill_viridis_c(name = "Prop. Single Mother", option = "viridis", direction = -1) +
  theme_tufte()  +
  theme(axis.text.x = element_text(angle = 90),
        legend.position = "bottom")
